Adapting to Quasi-Periodic and Secular Changes of Climate Risk

### 2018-03-03 writeup/main.tex
Dave updated methods

We generate synthetic flood sequences from two different simple conceptual models that represent chaotic dynamics and long memory.
For both cases, the expected value of the annual maximum flood flood follows a log-normal distribution, conditional on an expected value $\mu(t)$ which depends on individual model:
\begin{equation} \label{eq:lognormal}
  \log Q(t) \big| \mu(t),\sigma(t) \sim \normal \qty(\mu(t), \sigma(t)).
\end{equation}.
For simplicitiy we assume a constant coefficient of variation:
\begin{equation}\label{eq:calc-sigma}
  \sigma(t) = 
  \begin{cases} 
    \alpha \mu(t) & \qqtext{if} \alpha \mu(t) \geq \sigma_\text{min} \\ 
    \sigma_\text{min} & \qqtext{else}
  \end{cases}
\end{equation}
where $\sigma_\text{min}$ is a small constant that ensures non-trivial noise in the system.

@@

ENSO is the dominant mode of tropical climate variability, impacts flood risk around the world, and varies in a chaotic, quasi-oscillatory manner with variability on the order of 3 to 7 years with further low-frequency modulation \citep{Sarachik2010}.
A \SI{100000}{year} integration of the Cane-Zebiak model \citep{Zebiak1987} was used to produce a monthly NINO3.4 index with stationary forcing as described in \citet{Ramesh2017}.
To create an annual time series, we average the October-December values of the NINO3 index for each year.
A \SI{1000}{year} subset of this data is plotted in \cref{fig:enso-ts} along with the average spectral power of the series.

@@

This paper does not propose new models or suggest a general ``best'' model; instead it aims to describe the total uncertainty in estimates using easily methods which have been well studied in the literature.

@@

We use the log-normal model (\cref{eq:lognormal}) for its simplicity, interpretability, and good performance relative to other 2-parameter models \citep{Vogel1996}.
For each simulation, the first $N$ years \(\qty(t=[t_0-N+1, \ldots, t_0])\) are treated as observations (denoted \(Q_{\text{hist}}\) for clarity).
Then, the parameters are fit in a Bayesian framework to fully represent the posterior uncertainty as implemented in the \texttt{stan} probabilistic programming language \citep{Carpenter2016}.
Weak (uninformative) priors are chosen; in the context of a synthetic data experiment using informative priors seems inappropriate but other studies\footnote{cite} have used scaling information and regional information to inform prior distributions and reduce posterior variance.
The full probability model is thus:
\begin{align}
  \log Q_\text{hist} & \sim \normal \qty(\mu, \sigma) \\
  \mu &\sim \normal \qty(0, 10) \\
  \sigma &\sim \normal \qty(0, 2)
\end{align}

>>>>

We generate synthetic flood sequences from two different simple conceptual models that represent chaotic dynamics and long memory.
For both cases, the expected value of the annual maximum flood follows a log-normal distribution, conditional on an expected value $\mu(t)$, which in turn depends on individual models described in \cref{sec:methods-nino,sec:methods-markov}:
\begin{equation} \label{eq:lognormal}
  \log Q(t) \big| \mu(t),\sigma(t) \sim \normal \qty(\mu(t), \sigma(t)).
\end{equation}
For simplicity we assume a constant coefficient of variation:
\begin{equation}\label{eq:calc-sigma}
  \sigma(t) = 
  \begin{cases} 
    \alpha \mu(t) & \qqtext{if} \alpha \mu(t) \geq \sigma_\text{min} \\ 
    \sigma_\text{min} & \qqtext{else}
  \end{cases}
\end{equation}
where $\sigma_\text{min}$ is a small constant that ensures non-trivial noise in the system.

@@

ENSO is the dominant mode of tropical climate variability, impacts flood risk around the world, and varies in a chaotic, quasi-oscillatory manner with variability on the order of 3 to 7 years with further low-frequency modulation \citep{Sarachik2010}.
A \SI{100000}{year} integration of the Cane-Zebiak model \citep{Zebiak1987} was used to produce a monthly NINO3.4 index with stationary forcing as described in \citet{Ramesh2017}.
To create an annual time series, we average the October-December values of the NINO3 index for each year\footnote{So is this nino3.4 or 3?}.
A \SI{1000}{year} subset of this data is plotted in \cref{fig:enso-ts} along with the average spectral power of the series.

@@

This paper does not propose new models or suggest a general ``best'' model; instead it aims to describe the total uncertainty in estimates using methods that have been well studied in the literature.

@@

We use the log-normal model (\cref{eq:lognormal}) for its simplicity, interpretability, and good performance relative to other 2-parameter models \citep{Vogel1996}.
For each simulation, the first $N$ years \(\qty(t=[t_0-N+1, \ldots, t_0])\) are treated as observations (denoted \(Q_{\text{hist}}\) for clarity).
Then, the parameters are fit in a Bayesian framework to fully represent the posterior uncertainty as implemented in the \texttt{stan} probabilistic programming language \citep{Carpenter2016}.
Weak (uninformative) priors are chosen; in the context of a synthetic data experiment using informative priors seems inappropriate but other studies\footnote{cite} have used scaling information and regional information to inform prior distributions and reduce posterior variance\footnote{Prob you can drop this aside -- might just serve as a distraction to the main point here}.
The full probability model is thus:
\begin{align}
  \log Q_\text{hist} & \sim \normal \qty(\mu, \sigma) \\
  \mu &\sim \normal \qty(0, 10) \\
  \sigma &\sim \normal \qty(0, 2)
\end{align}
